Leadership across cultures and contexts: an exploratory analysis

### 2020-05-03 leadership_across_cultures_contexts.Rmd
Tightening up introduction a bit.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite the universality of the phenomenon of leadership, the function leaders serve across populations and social contexts is highly diverse and are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often take on essential roles in within-group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who fail to appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions carried out by a leader.

@glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

@@

<!-- WEIRD ORGNIZATIONAL STRUCTURE TELLS FUNCTIONS -->
In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service, nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders, to name a few [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions. The structure of such organizations are generally hierarchical with nested bodies and an internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

<!-- THEORY ON WEIRD FUNCTIONS -->
Scholars in managerial studies have long discussed the behaviors of leaders and managers have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [see @bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and often overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, create structures for accomplishing group goals, defend institutional integrity, and maintain internal consistency among group members.

<!-- DATA ON WEIRD FUNCTIONS -->
@wofford_factor_1970 provides an early empirical view of managerial functions. A factor analysis of managerial behaviors from over 130 US employees' survey responses identified five interpretable, orthogonal factors: (1) an *order and group achievement* factor, identifying participative, democratic, and organizational planning function; (2) a *personal enhancement oriented* factor, associated with authority and control, (3) a *personal interaction* factor, associated with interpersonal relationships and employee support, (4) a *security and maintenance* factor, associated with self-preservation, and (5) a *dynamic and achievement oriented* factor, associated with aggressive, self-assured actions. Wofford's analysis provided a foundation for future empirical studies of leadership and management functions. He contrasted factor one type management, described as the "professional administrator" who effectively collaborates and integrates team members, from factors two and five, described as "autocratic leaders" strongly motivated by personal power and a willingness to rely on aggressive tactics [@wofford_factor_1970].

<!-- COMPARATIVE DATA ON WEIRD FUNCTIONS -->
More recently, many managerial scholars have adopted integrative and comparative approaches to better understand cultural diversity in leadership systems. @behrendt_integrative_2017 draw on empirical results from the managerial literature and suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more fine-grained and externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. More integrated and holistic models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning. Culturally Endorsed Implicit Leadership Theory (CLT), suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied industries across 24 countries [@house2013strategic]. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.

The GLOBE CEO Study defines six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, the GLOBE CEO Study found that charismatic leadership behavior is consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and humane-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEOs behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By higher-order features of leadership behavior and situating leadership functions within more detailed classification frameworks, insights from managerial studies can more effectively be coupled with approaches from within psychology and anthropology.

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-industrial, and more traditional societies (i.e., non-WEIRD), from larger-scale, (post)industrial contexts (i.e., WEIRD), community leaders in non-WEIRD settings are more likely to have domain-specific functions based on role-specialization [see @smith_leadership_2016]. The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies. In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple domains. In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations. Community leadership functions in increasingly WEIRD contexts will likely be increasingly focused on economics and inter-group relations, whereas in less WEIRD contexts community leaders will primarily function to resolve interpersonal conflicts and manage intra-group social dynamics.
<!-- Could add ethnographic examples for SSS and LSS  -->

@@

<!-- UNIVERSAL QUALITIES OF COMMUNITY LEADERS -->
There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary discussion and consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 and conceptualize social status as a relative indicator of an individuals social value by the group and their access to contested resources within the group. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [@von_rueden_leadership_2015; @van_vugt_genes_2020], however, if there is causality between the two it is likely to be complex, context-specific, and driven by underlying correlates.

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. Empirical ethnographic studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. At least two studies relying on ethnographic fieldwork comparing leaders and non-leaders in relatively small-scale societies revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Across cultural contexts elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers in businesses across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as chiefdoms and many agricultural populations, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites, who maintain influence over community decisions are endowed with a high degree of social status rooted in ideologies and associated with their roles or kin group [@stanish_evolution_2010]. In such contexts community leaders often lack strong coercive authority, but maintain substantial influence through ideological systems and ritual or religious prowess inaccessible to non-elites [@stanish_evolution_2010]. The association of formal community leadership and supernatural abilities is likely an important precursor of institutionalized and coercive authority. Within state-level societies including non-industrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social and political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019]. 
<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formalized institutionalized roles and informal ephemeral roles, are also often if not universally associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but leadership roles through their association with increased social status are widely associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. 
<!-- BENEFITS: SOCIAL BENEFITS, SOCIAL NETWORK -->
Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].
<!-- NEED BETTER CITATIONS ABOVE -->
<!-- BENFITS: MATING -->
Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel meta-analysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

@@

@@

# Concluding remarks

>>>>

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite universality of the phenomenon of leadership, the function leaders serve across populations and social contexts are highly diverse and expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

Theoretical literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often take on essential roles in within-group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who fail to appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions of leaders.

@glowacki_leadership_2015 draw on empirical and ethnographic results in suggesting institutionalized leadership roles among small-scale societies are likely to initially emerge to resolve conflicts between families. Similarly, @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

@@

<!-- WEIRD ORGNIZATIONAL STRUCTURE TELLS FUNCTIONS -->
In WEIRD societies, community and political leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service, nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders, to name a few [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions. The structure of such organizations are generally hierarchical with nested bodies and an internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

<!-- THEORY ON WEIRD FUNCTIONS -->
Scholars in managerial studies have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [see @bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and often overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leaders functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, create structures for accomplishing group goals, defend institutional integrity, and maintain internal consistency among group members.

<!-- DATA ON WEIRD FUNCTIONS -->
@wofford_factor_1970 provides an early empirical view of managerial functions in a WEIRD setting. A factor analysis of managerial behaviors from over 130 US employees' survey responses identified five interpretable, orthogonal factors: (1) an *order and group achievement* factor, identifying participative, democratic, and organizational planning function; (2) a *personal enhancement oriented* factor, associated with authority and control, (3) a *personal interaction* factor, associated with interpersonal relationships and employee support, (4) a *security and maintenance* factor, associated with self-preservation, and (5) a *dynamic and achievement oriented* factor, associated with aggressive, self-assured actions. Wofford's analysis provided a foundation for future empirical work on management. He contrasted factor one type management, described as the "professional administrator" who effectively collaborates and integrates team members, from factors two and five, described as "autocratic leaders" strongly motivated by personal power and a willingness to rely on aggressive tactics [@wofford_factor_1970].

<!-- COMPARATIVE DATA ON WEIRD FUNCTIONS -->
More recently, many managerial scholars have adopted integrative and comparative approaches to better understand cultural or cross-national diversity in leadership systems. @behrendt_integrative_2017 draw on empirical results from the managerial literature and suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more fine-grained and externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. More integrated and holistic models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning. Culturally Endorsed Implicit Leadership Theory (CLT), suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied industries across 24 countries [@house2013strategic]. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, @house2013strategic suggest leaders moderate their behavior through culturally learned values and leadership-specific values.

The GLOBE CEO Study defined six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, they found that charismatic leadership behavior was consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and humane-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEO's behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By developing higher-order features of leadership behavior and situating leadership functions within more detailed classification frameworks, insights from managerial studies can more effectively be coupled with approaches from acroos behavioral science.

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies. In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexs. In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations. Community leadership functions in WEIRDer populations will likely be increasingly focused on economic contexts, creatign and managing organizational structures, and inter-group relations, whereas in less WEIRD contexts community leaders will primarily function to resolve interpersonal conflicts and manage intra-group social dynamics.
<!-- Could add ethnographic examples for SSS and LSS  -->

@@

<!-- UNIVERSAL QUALITIES OF COMMUNITY LEADERS -->
There are some broad commonalities in the qualities of community leaders across diverse populations. When strong community leaders emerge, they are most typically high status males who embody the qualities most respected by the group and necessary for successful leadership in the contexts in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 (and others) in conceptualizing social status as a relative indicator of the social value of the individual by the group and of their relative access to contested resources. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014], however, if there is a causal relationship it is likely complex, context-specific, and driven by underlying correlates.

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. @yanca_female_2004 found that sex-differences in political influence was more likely driven by ecological and social factors, rather than sex (or gender identity) per se, including social networks and access to same-sex allies. Two recent anthropological studies relying on ethnographic fieldwork among small-scale societies and comparing leaders and non-leaders, revealed sex is not a strong predictor of community influence or elected leadership, after accounting for economic specializations and a broad range of demographic, phenotypic, cognitive, and social traits [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Across cultural contexts elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers in businesses across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as chiefdoms and many agricultural populations, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status rooted in ideologies and associated with their roles or kin group [@stanish_evolution_2010]. In such contexts community leaders often lack strong coercive authority, but maintain influence through ideological systems and ritual or religious prowess inaccessible to non-elites [@stanish_evolution_2010]. The association of formal community leadership and supernatural abilities is likely an important precursor of institutionalized and coercive authority. Within state-level societies including nonindustrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social and political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019]. 
<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formalized institutionalized roles and informal ephemeral roles, are also often if not universally associated with increased social status (as previously discussed). Leadership roles, through their association with increased social status, are widely associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].
<!-- NEED BETTER CITATIONS ABOVE -->

<!-- BENFITS: MATING -->
Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel meta-analysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

@@

<!-- CITE OF CVR EMPIRICAL PAPER NEED TO DO -->

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
